Motion planning for assessing accessibility of protein tunnels

### 2017-09-10 paper.tex
experiments text updated, graphProbe regenerated

The proposed approach was used to analyze tunnel traversability towards the active site of Haloalkane dehalogenase protein (PDB ID 1CQW).
The active site was defined using residues 38 (Asparagine), 102 (Histidine) and 103 (Aspartate).
The molecular dynamics of the protein was computed using the Amber tool.
%100 frames of the dynamics were analyzed with the aim to calculate percentage of frames where the active site was accessible through the tunnels.
In each frame, tunnels for the spherical probe 0.9~\AA\ were detected using CAVER 3.0~\cite{caver3}. 
In most of the frames, two tunnels were detected.~\footnote{%
We refer to~\url{http://mrs.felk.cvut.cz/icra2018protein} for additional images.
}
The average bottleneck of the tunnels are $1.48$~\AA\ and $1.2$~\AA\ for the first and the second tunnel, resp.

Four different scaling-down factors were used: $\S=\{0.5,0.6,0.7,0.8\}$.
For each scale, 20 collision-free initial configurations were found in the $\RI=2$~\AA\ radius around the first sphere of the tunnel (outside the protein).
For each ligand, each minimal scale $\smin$ and each initial configuration, 10 trajectories were computed.

Two motion planners were utilized: the proposed planner (described in Alg.~\ref{alg::main} and Alg.~\ref{alg::expand}), 
    denoted as \RA\ in the rest of the paper, and the Retraction-based RRT~\cite{zhangRetraction} denoted as \RB\ in the rest of the paper.
The \RA\ method was run with the following parameters:
$\Imax=5,000$, $m=100$, $\gb=0.9$, $\dt=1.5$~\AA, $\rv=2$~\AA.
Both methods were implemented in the same software framework and they utilized the same data structures and routines for collision detection and nearest-neighbor search.
The nearest-neighbor search of both planners is based on the 6D Euclidean metric with the same weights for rotation and translation parts.
This metric was selected based on preliminary experiments with both planners as it resulted in the higher success rates.

The tested methods differ in the expansion step (lines \ref{alg::expa}--\ref{alg::expb} in Alg.~\ref{alg::expand}).
In these lines, the \RB\ method employs the retraction step in order to find a new collision-free configurations $\qnew$ that approaches $\qrand$.
The retraction step was realized by sampling the vicinity of the configuration $\qnear$ in the distance $0.1$~\AA\ using 20 samples.
A collision-free sample that maximally approached $\qrand$ was selected and added to the tree, and the retraction continued from this sample.
The retraction step was repeated 5 times.
The \RB\ method can therefore expand the tree by at most 5 new nodes per conformation, while \RA\ extends the tree only by one configuration.
The larger number of nodes in \RB\ method is not problematic, as the KD-trees were used for nearest-neighbor search and the time to solve these queries is significantly less than computing collision detection.
Parameters of the methods were selected in such a way that the number of collision detection queries in each expansion step is same.
For each conformation and tested scale, the \RA\ methods attempts to find $m=100$ collision-free configurations, and \RB\ attempts to perform $5 \times 20$ collision-detection queries.
Other parameters of the \RB\ planner were same as for \RA.
Planning was terminated after $\Imax$ iterations, or if the tree approached the active site to less than 0.5~\AA.

In the following experiments, we measure performance of the methods using the traversability rate, which is the percentage of the protein dynamic frames that are traversable.
A frame is considered traversable, if at least one trajectory out of 20 (the number of initial configurations $|\QI|$) $\times$ 10 trials approaches the active site to the distance less than 2~\AA.

@@

The results are shown in Tab.~\ref{tab::one} which shows traversability rate. 
The maximum percentage of traversable frames (the row Maximum in Tab.~\ref{tab::one}) is computed as the percentage of tunnels with bottleneck larger than the tested radius.
The row Planners shows traversable rate detected by the planners.
This rate decreases with the increasing radius and it is lower than the maximum traversable rate.
Higher traversability rate can be achieved by increasing the maximum number of planning iterations $\Imax$, which would also
increases the planning time.
For the purpose of the following experiments, we used $\Imax=5,000$ as a compromise between the performance and the runtime.
The rates provided by \RA\ and \RB\ are similar in this case.

@@

The traversability rates are shown in Tab.~\ref{tab::selection}, where the subscript of each ligand
name shows the number of selected conformations (5 or 50) and the used method (S or P).
The number after the slash in the ligand's name denotes the number of atoms of the ligand.
In the case of both tested ligands and 50 selected conformations, the `S' strategy leads to higher traversability rate than the `P' strategy.
For the ligand DCP, a higher traversability rate was achieved when 50 conformations were selected.
This experiments shows that selecting 50 conformations using the `S' strategy (Sec.~\ref{sec::strat}) leads to better results
than using less number of conformations or when preferring the packed conformations.

@@

The results are summarized in Tab.~\ref{tab::main}.
The tables shows percentages of traversable MD snapshots only for the first tunnel, as no trajectories were found in the second tunnel
for most of the tested ligand.
The only exception is the ligand m037t, which was the smallest tested ligand (with 8 atoms), and therefore it can fit even inside the second tunnel.
The traversability rate is higher for the most scaled-down ligands ($\smin=0.5$) and it decreases with the increasing scale.
For the scale $\smin=0.8$, the most traversable frames were detected for ligands DCP (8~\%), m037t (20~\%) and m038t (16~\%).

{\bf doufam, ze toto neni moc drze:}
The traversability of the DCP ligand to active site was approved by the MD simulation.
The observed trajectory of the DCP ligand followed the first tunnel.
From these results, we expect that at least ligands m037t and m038t can also reach the active site, but the MD simulations
haven't been performed yet.

>>>>

The proposed approach was used to analyze tunnel traversability towards the active site of Haloalkane dehalogenase protein (PDB ID 1CQW).
The active site was defined using residues 38 (Asparagine), 102 (Histidine) and 103 (Aspartate).
The molecular dynamics of the protein was computed using the Amber tool.
In each frame, tunnels for the spherical probe 0.9~\AA\ were detected using CAVER 3.0~\cite{caver3}. 
In most of the frames, two tunnels were detected.
The average bottleneck of the tunnels are $1.48$~\AA\ and $1.2$~\AA\ for the first and the second tunnel, resp.
The tunnel traversability was analyzed in 100 frames.

Four different scaling-down factors were used: $\S=\{0.5,0.6,0.7,0.8\}$.
For each scale, $|\QI|=20$ collision-free initial configurations were found in the $\RI=2$~\AA\ radius around the first sphere of the tunnel (outside the protein).
For each frame, minimal scale $\smin$ and each initial configuration, 10 trajectories were computed, which resulted
in $100 \times 4 \times 20 \times 10 = 80,000$ trajectories per ligand.

Two motion planners were utilized: the proposed planner (described in Alg.~\ref{alg::main} and Alg.~\ref{alg::expand}), 
    denoted as \RA\ in the rest of the paper, and the Retraction-based RRT~\cite{zhangRetraction} denoted as \RB\ in the rest of the paper.
The \RA\ method was run with the following parameters:
$\Imax=5,000$, $m=100$, $\gb=0.9$, $\dt=1.5$~\AA, $\rv=3$~\AA.
Both methods were implemented in the same software framework and they utilized the same data structures and routines for collision detection and nearest-neighbor search.
The nearest-neighbor search of both planners is based on the 6D Euclidean metric with the same weights for rotation and translation parts.
This metric was selected based on preliminary experiments with both planners as it resulted in the higher success rates.

The tested methods differ in the expansion step (lines \ref{alg::expa}--\ref{alg::expb} in Alg.~\ref{alg::expand}).
In these lines, the \RB\ method employs the retraction step in order to find a new collision-free configurations $\qnew$ that approaches $\qrand$.
The retraction step is realized by sampling the vicinity of the configuration $\qnear$ in the distance $0.1$~\AA\ using 20 samples.
A collision-free sample that maximally approaches $\qrand$ is selected and added to the tree, and the retraction continues from this sample.
The retraction step is repeated 5 times.
The \RB\ method can therefore expand the tree by at most 5 new nodes per conformation, while \RA\ extends the tree only by one configuration.
The larger number of nodes in \RB\ method is not problematic, as the KD-trees were used for nearest-neighbor search and the time to solve these queries is significantly less than computing collision detection.
Parameters of the methods were selected in such a way that the number of collision detection queries in each expansion step is same.
For each conformation and tested scale, the \RA\ methods attempts to find $m=100$ collision-free configurations, and \RB\ attempts to perform $5 \times 20$ collision-detection queries.
Other parameters of the \RB\ planner were same as for \RA.
Planning was terminated after $\Imax$ iterations, or if the tree approached the active site to less than 0.5~\AA.

In the following experiments, we measure performance of the methods using the traversability rate, which is the percentage of frames of the protein dynamic that are traversable by the tested ligand.
A frame is considered traversable, if at least one trajectory out of 20 (the number of initial configurations $|\QI|$) $\times$ 10 trials approaches the active site to the distance less than 2~\AA.~\footnote{We refer to~\url{http://mrs.felk.cvut.cz/icra2018protein} for additional material}

@@

The results are shown in Tab.~\ref{tab::one} which shows traversability rate. 
The maximum percentage of traversable frames (the row Maximum in Tab.~\ref{tab::one}) is computed as the percentage of tunnels with bottleneck larger than the tested radius.
The row Planners shows traversable rate detected by the planners.
This rate decreases with the increasing radius and it is lower than the maximum traversable rate.
The rates provided by \RA\ and \RB\ are similar which is a good indicator that the methods have been programmed correctly.
Higher traversability rate can be achieved by increasing the maximum number of planning iterations $\Imax$, which would also
increase the planning time.
For the purpose of the following experiments, we used $\Imax=5,000$ as a compromise between the performance and the runtime.

@@

The traversability rates are shown in Tab.~\ref{tab::selection}, where the subscript of each ligand's 
name shows the number of selected conformations (5 or 50) and the used method (S or P).
The number after the slash in the ligand's name denotes the number of atoms of the ligand.
In the case of both tested ligands and 50 selected conformations, the `S' strategy leads to higher traversability rate than the `P' strategy.
For the ligand DCP, a higher traversability rate was achieved when 50 conformations were selected.
This experiments shows that selecting 50 conformations using the `S' strategy (Sec.~\ref{sec::strat}) leads to better results
than using less number of conformations or when preferring the packed conformations.

@@

The results are summarized in Tab.~\ref{tab::main}.
The table shows the traversability rate only for the first tunnel, as no trajectory was found in the second tunnel for most of the tested ligands.
The only exception is the ligand m037t, which was the smallest tested ligand (with 8 atoms), and therefore it can fit even inside the second tunnel.
The traversability rate is higher for the most scaled-down ligands ($\smin=0.5$) and it decreases with the increasing scale.
For the scale $\smin=0.8$, the most traversable frames were detected for ligands DCP (8~\%), m037t (20~\%) and m038t (16~\%).

{\bf doufam, ze toto neni moc drze:}
The traversability of the DCP ligand to active site was approved by the MD simulation.
The observed trajectory of the DCP ligand in the MD simulation followed the first tunnel.
Based on the traversability rates with $\smin=0.8$, we expect that at least ligands m037t and m038t can also reach the active site, as
they have same (or even larger) traversability rate than DCP.
However, this hypothesis hasn't been tested agains MD simulations yet.
